---
title: "Untitled"
author: "Marcin Kierczak"
date: "12/4/2020"
output: html_document
---

```{r setup, include=FALSE}
local({
  hook_inline <- knitr::knit_hooks$get('inline')  # save the old hook
  knitr::knit_hooks$set(inline = function(x, options) {
    #x <- stringr::str_replace_all(x, 'ยง(.*?)ยง', '`r emo::ji(\'\\1\')`')
    x <- stringr::str_replace_all(x, 'em:(.*?)', 'emo::ji(\'\\1\')')
    hook_inline(x, options)
  })
})
knitr::opts_chunk$set(echo = TRUE)
```
## R Markdown

This is an R Markdown `r em:zebra` document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
